---
title: 'Are Roger and Serena the all-time best tennis players?'
author: 'Your Teacher Today: Bobby Stuijfzand'
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Before we begin

### Some steps to follow if you want to code along

1. Make sure RStudio is open
2. Create a new RStudio project in a new directory (name does not matter)
3. Create a folder in the RStudio project directory called `data`
4. Inside the main folder (not inside `data`), add the learner Rmarkdown file of this learner lab. You can find it here:   

On top of what you have installed for FDS, there are some additional packages that you need.
Don't forget to run the following chunk in the console before we begin.

```{r eval=FALSE}
install.packages("here")
install.packages("vroom")
```

### Beginning our R code

Load packages

```{r}
library(here)
library(tidyverse)
library(vroom)
```

## Defining the question

In this webinar we will use data from the WTP and ATP to answer the question:

***Who are the best tennis players of all-time?***

### The scope

We will limit this question to each gender separately - we are not going to compare between genders.

We will further look only at who won the grand slam single tournaments.


### Sub questions

We will use two questions to get some insight into our main question.

- Who won the most grand slam tournaments (single player)?
- Who won the most grand slam tournaments consecutively?

After we have found answer to these questions, we can combine this information to give a comprehensive answer to our main question.

## The data

### Credit where credit is due

The dataset we will be using is carefully assembled by [Jeff Sackmann](https://github.com/JeffSackmann) and is freely available on the internet in the form of github repositories, links for which you can find here:

- [https://github.com/JeffSackmann/tennis_atp](https://github.com/JeffSackmann/tennis_atp)
- [https://github.com/JeffSackmann/tennis_wta](https://github.com/JeffSackmann/tennis_wta)

To code along with this webinar you will have to download these repositories to your laptop and store them as folders under the `data` folder in your RStudio project. I will show you how to do this in the webinar.

### Exploring the data

Have a look at the repositories.

### Load the data

How do you load many files at once?

#### Get the filenames

We start with listing the files we are interested in with `list.files()`.

```{r}
data_paths_wta <- list.files(
  path = here("data", "tennis_wta-master"), 
  pattern = "^wta_matches_\\d.+csv"
)

data_paths_atp <- list.files(
  path = here("data", "tennis_atp-master"), 
  pattern = "^atp_matches_\\d.+csv"
)
```

#### Get the data itself

We will use `vroom()` to load in many datafiles with the same datastructure at once.

```{r}
data_wta <- vroom::vroom(
  here::here("data", "tennis_wta-master", data_paths_wta), 
  id = "file"
) 

data_atp <- vroom::vroom(
  here::here("data", "tennis_atp-master", data_paths_atp), 
  id = "file"
) 
```

Let's now admire our data:

```{r}
head(data_wta) %>% knitr::kable()
```

It's a bit wide thanks to the `file` column's content, but using `kable()` makes that we still have a nice format to explore this data, though I wouldn't suggest keeping it this way in a final report.

--- 

#### EXERCISE

Let's look at some columns we might be interested in using `select()`.
What columns might you include in your report to communicate a general sense about this data?

```{r}
data_wta %>% 
  sample_n(20) %>% 
  select( ####fill in some column names here#### ) %>% 
  knitr::kable()
```
---

## Preparing the data

### Creating the relevant subset 

We want to only keep Grand Slam data, and only data on the final round.

---

#### EXERCISE 

Using the columns `tourney_level` and `round`, can you work out how to filter the data so that we only keep a subset that satisfies the conditions above?

```{r, cache=TRUE}
data_wta_g_winners <- data_wta %>% 
  filter(tourney_level == .., round == ..)

data_atp_g_winners <- data_atp %>% 
  filter(tourney_level == .., round == ..)
```

---

### Adding dates

Change the `tourney_date` column into a date type and sort our data.

```{r}
data_wta_g_winners <- data_wta_g_winners %>% 
  mutate(date = lubridate::ymd(tourney_date)) %>% 
  arrange(date)

data_atp_g_winners <- data_atp_g_winners %>% 
  mutate(date = lubridate::ymd(tourney_date)) %>% 
  arrange(date)
```

Our data is now ready.

## First question: Who won the most grand slam tournaments (single player)?

---

#### EXERCISE

What code would you use to get the top 5 winners for each gender?

```{r}
data_wta_g_winners %>% 
  count(winner_name, sort = TRUE) %>% 
  top_n(n, n =) %>% 
  knitr::kable(caption = "Top 5 Grand Slam Women",
               col.names = c("Athlete", "*n*"))
```

```{r}
data_atp_g_winners %>% 
  count(winner_name, sort = TRUE) %>% 
  top_n(n, n =) %>% 
  knitr::kable(caption = "Top 5 Grand Slam Men",
               col.names = c("Athlete", "*n*"))
```

---

## Second question: Who won the most grand slam tournaments consecutively?

We need to keep the following things in mind:

- An athlete can win once, then lose, and then win again. We don't want these two wins to count towards the same streak.
- An athlete can have multiple streaks, so the answer will not be as unambiguous as in the previous part.

I will demonstrate an approach in steps for the women.
Then you can use this as an exercise to solve it for the men.

### Step 1: Number each game consecutively

We will use `row_number()` for this.

```{r}
data_wta_g_cons_winners <- data_wta_g_winners %>% 
  mutate(game_id = row_number()) 
```

We now have a number for each game:

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  head(10) %>% 
  select(tourney_name, date, game_id) %>% 
  View()
```

### Step 2: Group our data by athlete

```{r}
data_wta_g_cons_winners <- data_wta_g_cons_winners %>% 
  group_by(winner_name)
```

Before we continue, let's have a look at the data.

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  arrange(winner_name) %>% 
  select(winner_name, game_id, date) %>% 
  View()
```

### Step 3: Find the id of the previous game_id

We can use `lag()` to find the previous `game_id` an athlete has won.
We use `default = -99` so that if there is no previous game, we still have a value to work with.

```{r}
data_wta_g_cons_winners <- data_wta_g_cons_winners %>% 
  mutate(previous_game = lag(game_id, default = -99))
```

Let's examine the data again:

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  arrange(winner_name) %>% 
  select(winner_name, game_id, previous_game, date) %>% 
  View()
```

### Step 4: Is the current id the start of something new?

```{r}
data_wta_g_cons_winners <- data_wta_g_cons_winners %>% 
  mutate(streak_begin = game_id - previous_game != 1)
```

Let's examine the data again:

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  arrange(winner_name) %>% 
  select(winner_name, game_id, previous_game, streak_begin) %>% 
  View()
```

### Intermezzo: the cumulative sum

For the next step, we are going to process the `streak_begin` variable in a way that we will get a `streak_id` for each separate streak.
In order to do so we can use a function called `cumsum()` which calculates the cumulative sum for a variable.

What is the cumulative sum?

Let's demonstrate.

We have a vector containing the values `1, 2, 3, 4`.

Calculating the cumulative sum gives:

```{r}
cumsum(c(1,2,3,4))
```

The cumulative sum returns a vector of the same length, that contains for each element the sum of the original vector *up until that point*.

How is this useful?

Consider that `TRUE` and `FALSE` are represented by `1` and `0` in R.
A cumulative sum on such a logical vector would look like this:

```{r}
cumsum(c(FALSE, TRUE, FALSE, FALSE, TRUE))
```

You can see that only once a `TRUE` appears will the cumulative sum increment.

Now recall that in our `streak_begin` variable, we have a `TRUE` for every start of a streak.

Let's look at the data again:

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  arrange(winner_name) %>% 
  select(winner_name, game_id, previous_game, streak_begin) %>% 
  View()
```

### Step 5: Get an id for each streak

If we run `cumsum()` on `streak_begin`, what we will get is a number that increments with every new beginning.

```{r}
data_wta_g_cons_winners <- data_wta_g_cons_winners %>% 
  mutate(streak_id = cumsum(streak_begin))
```

Let's admire our result.

```{r eval=FALSE}
data_wta_g_cons_winners %>% 
  arrange(winner_name) %>% 
  select(winner_name, game_id, previous_game, streak_begin, streak_id) %>% 
  View()
```

### Step 6: Let's count the streaks!

Now all that's left is counting up the streak ids to understand how many rows belonged to each streak, and therefore, how long each streak was!

```{r}
data_wta_g_cons_winners_result <- data_wta_g_cons_winners %>% 
  count(streak_id, sort=TRUE)
```

Let's admire the result

```{r}
data_wta_g_cons_winners_result %>% 
  head(20) %>% 
  knitr::kable()
```

As we stored the results in a separate tibble, We can now also go back to the original tibble to examine some individual streaks.

For example Serena Williams longest streak:

```{r}
data_wta_g_cons_winners %>% 
  filter(streak_id == 13) %>% 
  select(winner_name, date, tourney_name)
```

## And now for the men...

---

#### EXERCISE: 

Can you answer the second question for the men?

---

## Open era tennis results

---

#### BONUS:

What would the results look like if we account for the switch to allowing professionals to compete in grand slams?

Hint: the open era, as it was called, started in 1968.

---
